---
title: "Creating and validating a model"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Creating and validating a model}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

```{r setup, messages=FALSE}
library(NHSRtt)
library(dplyr)
library(tidyr)
library(lubridate)
```


This vignette provides a walkthrough for how to use the NHSRtt package.

The vignette shows:

1. how the data are set up
2. how the renege and capacity parameters are calculated
3. how the parameter performance can be validated
4. how the paramters can be applied to future scenarios

## Setting up the data

The public data can be sourced with the `get_rtt_data()` function.
Note, at the moment this provides one dataset, but this will be updated so the user specifies which data they will obtain.

Conceptually, we want to use a year of data to calibrate parameters, and then the following six months to validate those parameters.

Within this code, the `convert_months_waited_to_id()` is used to convert the string for months_waited into a numeric value.

```{r data_download}

calibration_start <- as.Date("2023-01-01")
calibration_end <- as.Date("2023-12-31")
validation_end <- as.Date("2024-06-01")
analysis_trust <- "RC9" # filter for a single trust
max_months_waited <- 12 # I am only interested in waiting time bins up to 12 months

monthly_rtt <- NHSRtt::get_rtt_data(
  date_start = calibration_start,
  date_end = validation_end,
  show_progress = FALSE # can change this to TRUE to see progress
) |>
  filter(
    trust == analysis_trust
  ) |> 
  mutate(
    type = case_when(
      type == "Non-Admitted" ~ "Complete",
      type == "Admitted" ~ "Complete",
      type == "New Periods" ~ "Referrals",
      .default = type
    ),
    months_waited_id = convert_months_waited_to_id(
      months_waited,
      max_months_waited
    )
  ) |>
  summarise(
    value = sum(value),
    .by = c(
      trust,
      period,
      type,
      months_waited_id
    )
  ) |> 
  arrange(
    type,
    months_waited_id,
    period
  ) |> 
  mutate(
    period_id = dplyr::row_number(), # we need period_id for later steps
    .by = c(
      type,
      months_waited_id
    )
  )

head(monthly_rtt)
```

The data need to be in a specific format (as described by the function documentation) for the functions to work.
To calculate the renege and capacity parameters, the inputs you needs are:

1. Referrals per time step (these are inputs to each time step), where all referrals have waited 0 months
2. Complete pathways per time step by the number of months waited
3. Incomplete pathways per time step by the number of months waited

```{r data_manipulation}

calibration_period <- monthly_rtt |>
  filter(
    between(
      period,
      calibration_start,
      calibration_end
    )
  ) |>
  summarise(
    value = sum(value),
    .by = c(
      trust,
      period_id,
      type,
      months_waited_id
    )
  )

referrals <- calibration_period |>
  filter(
    type == "Referrals"
  ) |>
  distinct(
    period_id,
    value
  ) |>
  rename(
    referrals = "value"
  )

print(referrals)

completes <- calibration_period |>
  filter(
    type == "Complete"
  ) |>
  distinct(
    period_id,
    months_waited_id,
    value
  ) |>
  rename(
    treatments = "value"
  )

print(completes)

incompletes <- calibration_period |>
  filter(
    type == "Incomplete"
  ) |>
  distinct(
    period_id,
    months_waited_id,
    value
  ) |>
  rename(
    incompletes = "value"
  )

print(incompletes)

```

## Generating the model parameters

These inputs can get passed into the `calibrate_capacity_renege_params()` function.

For occasions where renege counts are negative for the group of people that have been waiting 0 months, we can tell the modelling to assume these are under-reported referrals and redistribute these to referrals. This is done by setting the `redistribute_m0_reneges` to TRUE.

```{r generate-params}
params <- calibrate_capacity_renege_params(
  referrals = referrals,
  completes = completes,
  incompletes = incompletes,
  max_months_waited = max_months_waited,
  redistribute_m0_reneges = TRUE, 
  full_breakdown = FALSE # this can be set to TRUE to see all the transitions for all moths waited at each time step
)

print(params)
```


## Validate these parameters on a known time series

Pass the parameters to a known timeseries, and calculate an evaluation metric, comparing the predicted count of incomplete pathways with the known count of incomplete pathways.

The example below calculates the mean absolute percentage error (MAPE).

First, the data needs constructing.
The data required are:

1. a vector of capacity per time step
2. a vector of referrals per time step
3. a table of counts of incomplete pathways by months waiting for the time step prior to the first time step of the other two data inputs

```{r validation-data}

validation_period <- monthly_rtt |>
  filter(
    between(
      period,
      lubridate::floor_date(calibration_end, unit = "months"), # this helps the user pick up the incompletes from the previous month
      validation_end
    )
  ) |>
  summarise(
    value = sum(value),
    .by = c(
      trust,
      period_id,
      type,
      months_waited_id
    )
  )

validation_referrals <- validation_period |>
  filter(
    type == "Referrals",
    period_id != min(period_id)
  ) |>
  distinct(
    period_id,
    value
  ) |>
  arrange(
    period_id
  ) |> 
  pull(
    value
  )

print(validation_referrals)

validation_capacity <- validation_period |>
  filter(
    type == "Complete",
    period_id != min(period_id)
  ) |> 
  summarise(
    count = sum(value),
    .by = period_id
  ) |>
  arrange(
    period_id
  ) |> 
  pull(
    count
  )

print(validation_capacity)

incompletes_at_t0 <- validation_period |>
  filter(
    type == "Incomplete",
    period_id == min(period_id)
  ) |>
  distinct(
    months_waited_id,
    value
  ) |>
  rename(
    incompletes = "value"
  )

print(incompletes_at_t0)
```

The data are then passed to `apply_params_to_projections()` which provides a full table of stocks at each time step.

```{r validation}
validation_performance <- apply_params_to_projections(
  capacity_projections = validation_capacity,
  referrals_projections = validation_referrals,
  incomplete_pathways = incompletes_at_t0,
  renege_capacity_params = params,
  max_months_waited = max_months_waited
)

print(validation_performance)

```

MAPE is then calculated from this

```{r mape}
# calculate mean absolute percentage error
# 1. calculate observed incompetes by months waited and period
observed_incompletes <- validation_period |> 
  filter(
    type == "Incomplete",
    period_id != min(period_id)
  ) |> 
  summarise(
    observed = sum(value),
    .by = c(
      period_id, months_waited_id
    )
  )  |> 
  mutate(
    period_id = period_id - min(period_id) + 1 # to change the period_ids to the same scale as the ones calculated
  )

# 2. quantify observed incompletes by months waited and period
estimated_incompletes <- validation_performance |> 
  dplyr::select(
    "period_id", 
    "months_waited_id",
    predicted = "incompletes"
  )

# 3. calculate mape
observed_incompletes |> 
  left_join(
    estimated_incompletes,
    by = join_by(
      period_id, months_waited_id
    )
  ) |> 
  mutate(
    absolute_percentage_error = abs(
      (predicted - observed) / observed
    )
  ) |> 
  summarise(
    mean_absolute_percentage_error = mean(
      absolute_percentage_error
    )
  )

```

